Model matters with restricted Boltzmann machines

### 2018-05-08 presentations/msmlc/slides.Rmd
small typos

---
title: "Model matters with restricted Boltzmann machines"
shorttitle: "Model matters with RBMs"
author: Andee Kaplan
shortname: Kaplan, et al.
institute: |
    | Iowa State University
    | ajkaplan@iastate.edu
shortinstitute: ajkaplan@iastate.edu
date: |
  | May 9, 2017
  |
  | Slides available at <http://bit.ly/kaplan-msmlc>
  | 
  | \footnotesize Joint work with D. Nordman and S. Vardeman
shortdate: "May 9, 2017"
output: 
  beamer_presentation:
    template: ../resources/latex/beamer.tex
    includes:
      in_header: ../resources/latex/front-matter.tex
theme: CambridgeUS
bibliography: ../resources/latex/refs.bib
fig_caption: true
nocite: |
    @salakhutdinov2009deep, @kaiser2007statistical
---

@@

\begin{definition}[S-unstable RBM]
A RBM model formulation is \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_{\nv}) = \infty.
\end{align*}
where 
\begin{align}
\label{eqn:elpr}
\text{ELPR}(\boldsymbol \theta_{\nv}) = \log \left[\frac{\max\limits_{(v_1, \dots, v_{\nv}) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right]
\end{align}
\end{definition}

@@

\begin{result}
\label{prop:instab}
Let $c > 0$ and fix an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
\end{result}

If the $\nv^{-1}ELPR(\boldsymbol\theta_{\nv})$ is too large, then a RBM model will exhibit large probability shifts for very small changes in the data configuration.

@@

- To explore the effects of RBM parameters $\boldsymbol \theta$ on *near-degeneracy*, *instability*, and *uninterpretability*, consider models of small size  
- For $\nh, \nv \in \{1, \dots, 4\}$, sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$
    1. Split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$
    2. Allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$
    3. Average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points
- Calculate metrics of model impropriety, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$. 
- In the case of *near-degeneracy*, classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$

@@

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(frac(bar(ELPR(theta)), n[v])), low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom") -> p.elpr
```

```{r, fig.show="hold", fig.width=3, out.width='.32\\linewidth', fig.cap="\\label{fig:degen_plots}The fraction of models that were near-degenerate (left), the sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ (middle), and the sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ (right)."}
p.degen
p.elpr
p.exp_diff
```

@@

1. Generalization of instability results for other network models [ongoing, see @kaplan2016note]
2. Image classification
    - Ensemble methods (super learners) using AdaBoost [@freund1995desicion] 
    - Decision theoretic based approach to approximating the likelihood ratio test for classification
3. Markov chain Monte Carlo methods for data with Markovian dependence
    - Spatial data
    - Network data

@@

* Contact
    * Email -- <ajkaplan@iastate.edu>
    * Twitter -- <http://twitter.com/andeekaplan>
    * GitHub -- <http://github.com/andeek>

>>>>

---
title: "Model matters with restricted Boltzmann machines"
shorttitle: "Model matters with RBMs"
author: Andee Kaplan
shortname: Kaplan, et al.
institute: |
    | Duke University
    | andrea.kaplan@duke.edu
shortinstitute: andrea.kaplan@duke.edu
date: |
  | May 9, 2018
  |
  | Slides available at <http://bit.ly/kaplan-msmlc>
  | 
  | \footnotesize Joint work with D. Nordman and S. Vardeman
shortdate: "May 9, 2018"
output: 
  beamer_presentation:
    template: ../resources/latex/beamer.tex
    includes:
      in_header: ../resources/latex/front-matter.tex
theme: CambridgeUS
bibliography: ../resources/latex/refs.bib
fig_caption: true
nocite: |
    @salakhutdinov2009deep, @kaiser2007statistical
---

@@

\begin{definition}[S-unstable RBM]
A RBM model formulation is \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_{\nv}) = \infty.
\end{align*}
where 
\begin{align}
\label{eqn:elpr}
\text{LREP}(\boldsymbol \theta_{\nv}) = \log \left[\frac{\max\limits_{(v_1, \dots, v_{\nv}) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right]
\end{align}
\end{definition}

@@

\begin{result}
\label{prop:instab}
Let $c > 0$ and fix an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
\end{result}

If the $\nv^{-1}LREP(\boldsymbol\theta_{\nv})$ is too large, then a RBM model will exhibit large probability shifts for very small changes in the data configuration.

@@

- To explore the effects of RBM parameters $\boldsymbol \theta$ on *near-degeneracy*, *instability*, and *uninterpretability*, consider models of small size  
- For $\nh, \nv \in \{1, \dots, 4\}$, sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$
    1. Split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$
    2. Allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$
    3. Average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points
- Calculate metrics of model impropriety, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{LREP}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$. 
- In the case of *near-degeneracy*, classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$

@@

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(frac(bar(LREP(theta)), n[v])), low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom") -> p.elpr
```

```{r, fig.show="hold", fig.width=3, out.width='.32\\linewidth', fig.cap="\\label{fig:degen_plots}The fraction of models that were near-degenerate (left), the sample mean value of $\\text{LREP}(\\boldsymbol \\theta)/\\nv$ (middle), and the sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ (right)."}
p.degen
p.elpr
p.exp_diff
```

@@

1. Generalization of instability results for other network models [ongoing, see @kaplan2016note]
\vfill
2. Image classification
    - Ensemble methods (super learners) using AdaBoost [@freund1995desicion] 
    - Decision theoretic based approach to approximating the likelihood ratio test for classification
\vfill

@@

* Contact
    * Email -- <andrea.kaplan@duke.edu>
    * Twitter -- <http://twitter.com/andeekaplan>
    * GitHub -- <http://github.com/andeek>
